---
title: "CodeBook"
author: "Debajit Ghosh"
date: "July 25, 2015"
output: html_document
---
> Objective of the R Script

* To create a tidy data that could be used later for analysis

> Data Overview

* Human Activity Recognition database built from the recordings of 30 subjects performing activities of daily living (ADL) while carrying a waist-mounted smartphone with embedded inertial sensors.

  + Data Set Characteristics: Multivariate, Time-Series
  + Number of Instances: 10299
  + Number of Attributes: 561
  

> Data Collection

* The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years
* Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist
* Using its embedded accelerometer and gyroscope, 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz was captured
* The experiments have been video-recorded to label the data manually
* The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

> PART 1: Data Extraction 

* PART 1 of the R Script run_analysis -
  + Downloads the zip file and saves it as GACDataCProj.zip in the working directory
  + Unzips GACDataCProj.zip to the working directory
  + Directory "UCI HAR Dataset" contains all files need for running the R Script


```{r}
list.files("./UCI HAR Dataset")
```

> Data Sets

* features_info.txt: Shows information about the variables used on the feature vector
* features.txt : List of all features - contains the list of 561 features that needs to be mapped to the test and training datasets

```{r}
str (read.table("./UCI HAR Dataset/features.txt"))
```

* activity_labels.txt: Links the class labels with their activity name (6 activities)

```{r}
str (read.table("./UCI HAR Dataset/activity_labels.txt"))
(read.table("./UCI HAR Dataset/activity_labels.txt"))[,2]
```

* Test and training datasets can be found under the respective sub directories named "test" and "train"
  * Unit of Measurement: The features selected for this database come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. 
  + train/X_train.txt: Training set - A 561-feature vector with time(t) and frequency(f) domain variables
      ++ match column names to features.txt for descriptive labels
  + train/y_train.txt: Training labels - Its activity label
      ++ match with activity_labels.txt for descriptive labels 
  + train/subject_train.txt: subject who performed the test  
    
  + test/X_test.txt': Test set- A 561-feature vector with time and frequency domain variables
      ++ match column names to features.txt for descriptive labels
  + test/y_test.txt': Test labels - Its activity label
      ++ match with activity_labels.txt for descriptive labels
  + test/subject_test.txt: subject who performed the test  

```{r}
dim (read.table("./UCI HAR Dataset/test/X_test.txt"))
dim (read.table("./UCI HAR Dataset/test/y_test.txt"))
dim(read.table("./UCI HAR Dataset/test/subject_test.txt"))
dim (read.table("./UCI HAR Dataset/train/X_train.txt"))
dim (read.table("./UCI HAR Dataset/train/y_train.txt"))
dim(read.table("./UCI HAR Dataset/train/subject_train.txt"))
```

> PART 2: Joining tables to create a master dataset 

* PART 2 of the R script run_analysis joins the test and training datsets, adds column names, activity labels and subject list 
  * 2.1 creates a vector of column names from features.txt 
```{r, echo=FALSE}
dim (t(read.table("./UCI HAR Dataset/features.txt")))
```
  *2.2 and 2.3 creates a dataset named "test" and "train" respectively 
    ++ replaces column names in the X_train and X_test datsets with names from features.txt
    ++ adds the subject from subject_train and subject_test
    ++ adds the activity labels from y_train and y_train and references activity_labels.txt for more descriptive labels
  * 2.4 Appends the train dataset to the test dataset created in 2.2 and 2.3 to create "datamaster"
  
```{r, echo=FALSE}
  columns <- read.table("./UCI HAR Dataset/features.txt")
    columns1 <- t(columns)
      
## 2.2 Read the test data, add column names, activity labels and subject that performed the test 
    
    testdata <- read.table("./UCI HAR Dataset/test/X_test.txt")
      colnames(testdata) <- c(columns1[2,1:ncol(columns1)]) ## adds column names from 2.1
        testlabels <- read.table("./UCI HAR Dataset/test/y_test.txt")
          testsubject <- read.table("./UCI HAR Dataset/test/subject_test.txt")
            test <- cbind(testlabels,testsubject, testdata)
              names(test)[1] <- "Label" ## adding descriptive labels 
                names(test)[2] <- "Subject" ## adding descriptive labels
               
              
      
## 2.3 Read the train dataset, add column names, activity labels and subject that performed the test 
            
                traindata <- read.table("./UCI HAR Dataset/train/X_train.txt")
                  colnames(traindata) <- c(columns1[2,1:ncol(columns1)]) ## adds column names from 2.1
                    trainlabels <- read.table("./UCI HAR Dataset/train/y_train.txt")
                      trainsubject <- read.table("./UCI HAR Dataset/train/subject_train.txt")
                        train <- cbind(trainlabels,trainsubject, traindata)
                          names(train)[1] <- "Label" ## adding descriptive labels 
                            names(train)[2] <- "Subject" ## adding descriptive labels 
## 2.4 Join the test data from 2.2 and train data from 2.3
                      
                                datamaster <- rbind(test,train)
                                  dim(datamaster)
```

> PART 3: Extracts only the measurements on the mean and standard deviation for each measurement and adds descriptive activity names to name the activities in the data set

* Part 3 of the R Script run_analysis (Please Note: This portion of the script utilizes the dplyr package) - 
  * 3.1 Subsets the data master created in 2.4 based on the column names containing mean and standard deviation while retaining the Label and Subject columns
  * 3.2 Further subsets the dataset created in 3.1 to exclude columns containing Mean Frequencies which is a weighted average of the frequency components and not a straught arithmetic mean
  * 3.3 Reads the activity list table and changes the column names to match column names in the meansd and datamaster datasets
  * 3.4 Adds descriptive activity labels from activiylist to the meansd dataset to create the meansd_act dataset

```{r}
library(dplyr)
## Extracts only the measurements on the mean and standard deviation for each measurement and adds descriptive activity names to name the activities in the data set

   
## 3.1 Subsets the data master created in 2.4 based on the column names containing mean and standard deviation while retaining the Label and Subject columns
                  
                  meansd <- datamaster[,grep("mean()|std()|Label|Subject", names(datamaster))]

## 3.2 Further subsets the dataset created in 3.1 to exclude columns containing Mean Frequencies which is a weighted average of the frequency components and not a straught arithmetic mean
                  
                      meansd <- meansd[,grep("Freq", names(meansd), invert=TRUE)] 

## 3.3 Reads the activity list table and changes the column names to match column names in the meansd and datamaster datasets

                        activitylist <- read.table ("./UCI HAR Dataset/activity_labels.txt")
                          names(activitylist)[1] <- "Label"
                            names(activitylist)[2] <- "Activity"
                          
## 3.4 Adds descriptive activity labels from activiylist to the meansd dataset to create the meansd_act dataset 
                            
                              meansd_act <- activitylist %>%
                                            right_join(meansd, by="Label")                             
dim(meansd_act)
```

> PART 4: Labels the data set with descriptive variable names wherever approrpiate

* Part 4 of the R Script run_analysis -
  * Subsititutes specific character strings in the column names of the dataset meansd_act with more descriptive labels

```{r, echo=FALSE}
    sample <- meansd_act[1,5]
    head(sample)
```

> PART 5: Creates a second, independent tidy data set with the average of each variable for each activity and each subject

* Part 5 of the R script run_analysis does the following (Please Note: This portion the script requires the dplyr and tidyr packages) -
  * Creates a text file TidyData.txt in the workling directory that gives a grouped mean for all selected features by activity and subject 
  * The data is arranged in a neat tidy format in 7 columns and 11880 rows
```{r, echo=FALSE}
                                      library(tidyr)
                                      meansd_act %>%
                                          group_by(Activity,Label,Subject) %>%
                                                summarise_each(funs(mean)) %>%
                                                        arrange(Subject,Activity) %>%
                                                            gather("ActivityDescription-Stats-Axes","Value",4:ncol(meansd_act)) %>%
                                                                separate("ActivityDescription-Stats-Axes",c("ActivityDescription","Stats","Axes")) %>%
                                                                  dim()
```
  
